---
title: "General Introduction to ecositer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General Introduction to ecositer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is intended to document basic functionality of the *ecositer* package. This package is in development and should not be expected to perform in a robust manner across different datasets. At this point, the functionality here is a demonstration of workflows for dealing with NRCS ecological site data. I hope that future development of this package and standardization of NASIS population will make it more applicable. 

Library
```{r setup, message=FALSE, warning=FALSE}
library(ecositer)
library(soilDB)
library(dplyr)
library(indicspecies)
library(stringr)
library(listviewer)
library(plotly)
library(kableExtra)
```

Begin by querying NASIS so that you have a selected set containing vegetation data. In this vignette, we will query for vegetation data in Sequoia and Kings Canyon National Parks, soil survey area CA792. Detailed instructions for making this query can be found [here](https://natearoe.github.io/ESS_FAQs/ecosites.html#plotveg).

Now that we have data in our selected set, we will introduce the first *ecositer* function. ```vegStaticNASIS()``` takes a snapshot of your NASIS selected set and saves it in a local SQLite database. It is referred to as "static" because it does not change. This means that regardless of what is in your selected set in the future, you can refer to this static NASIS database using a local file path on your harddrive. Rather than switching between different queries and clearing your selected set in NASIS, you will likely find it useful to save a few specific selected sets locally on your computer. The functionality of creating static NASIS selected sets is borrowed from [soilDB](https://github.com/ncss-tech/soilDB). ```vegStaticNASIS()``` is a special case designed specifically for vegetation data. 

The only argument in this function is ```output_path``` the local file path where you want to save your static NASIS database. 
```{r, eval=FALSE}
ecositer::vegStaticNASIS(output_path = "C:/Users/Nathan.Roe/Documents/SEKI/CA792_veg.sqlite")
```

Let's start off by looking at how many pedons and vegplots we have for each ecosite
```{r, message=FALSE, warning=FALSE}
numb_plots <- number_plots_by_site(static_location = "C:/Users/Nathan.Roe/Documents/SEKI/CA792_veg.sqlite")
head(numb_plots)
```

Now we can format our vegetation data into a more useful format
```{r, message=FALSE, warning=FALSE}
CA792_veg_formatted <- formatted_veg_df(static_location = "C:/Users/Nathan.Roe/Documents/SEKI/CA792_veg.sqlite")
```

Here is a look at a portion of the data we have assembled.
```{r, warning=FALSE}
CA792_veg_formatted |> dplyr::select(-primarydatacollector, - plantnatvernm, -akfieldecositeid,
                                                   -vegplotid, -vegplotiid, -plantsym)
```

Now that the data is formatted we can begin to analyze it. 
```{r}
CA792_veg_summary <- ecositer::veg_summary(veg_df = CA792_veg_formatted)
```

The result of ```veg_summary()``` is a nested list. This list provides a vegetation summary for every ecosite and state/phase in your dataset, cover data by strata, as well as raw dataset by ecosite and state/phase. Below is an overview of the general structure of the results.  
```{r}
str(CA792_veg_summary[c(1,2,3)], max.level = 2)
```

Now let's look at a vegetation summary for the ecosite R022AB006CA. There are 15 columns with different attribute
```{r}
head(CA792_veg_summary$R022AB006CA$Cover_data)
```

One useful tool for looking at ecological data is Non-Metric Multidimensional Scaling. 
```{r, message=FALSE, warning=FALSE}
my_nmds <- ecositer::nmds_ecosite(veg_summary = CA792_veg_summary, ecosite = c("F022AK100CA", "F022AK101CA"), pres_abs = TRUE, nmds_dim = 2, reduce_species = NA)

```

Visualize NMDS
```{r, message=FALSE, warning=FALSE}
nmds_plot(static_location = "C:/Users/Nathan.Roe/Documents/SEKI/CA792_veg.sqlite", 
          nmds = my_nmds,
          veg_summary = CA792_veg_summary)
```


