---
title: "Vegetation data ordination"
author: "Nathan Roe"
date: "2025-01-02"
output: html_document
---

Ordination is a multivariate statistical technique that is well-suited for analyzing ecological communities. Ordination allows for many ecologically useful relationships to be evaluated - all correlative, of course (this is ecology after all) - including:
- similarity/dissimilarity in vegetative composition between sites
- relationship between community classifications and environmental gradients
- relationship between community classifications and species
- relationship between species and environmental gradients

This vignette is not intended to explain the mechanics of ordination techniques, distance measures, or how to interpret ordination results. There are many excellent sources for this purpose. I would recommend reading [Analysis of Ecological Communities by McCune and Grace](https://www.researchgate.net/publication/216769990_Analysis_of_Ecological_Communities) for a primer on these topics. 

In this vignette I will present the use of Non-metric Multidimensional Scaling (NMDS) with a Bray-Curtis distance measure for creating an ordination. Much of the workflow will not be built into `ecositer` functions because it relies heavily on functions from other packages, and incorporating them into functions will conceal too many of parameters that can be, and should be, manipulated in this process. 

I will be using the CA792 (Sequoia and Kings Canyon National Park) Soil Survey for this demonstration.

For this example, I will be working with pedon and vegetation data stored in the `ecositer.data` package.
```{r, results="hide", message=FALSE}
data(CA792_pedon_data, package = "ecositer.data")
data(CA792_veg_data, package = "ecositer.data")
```

```{r, echo=FALSE, results="hide", message=FALSE}
get("CA792_pedon_data")
get("CA792_veg_data")
```

We will produce environmental variables, as demonstrated in the [Summarize Ecological Data vignette](https://natearoe.github.io/ecositer/articles/summarizing_ecological_data.html). 
```{r}
CA792_pedon_summary <- ecositer::summarize_pedon_soil_properties(SS = FALSE,
                                           r_object = CA792_pedon_data,
                                           byDepth = list(c(0, 25), c(0, 50)))

```


