---
title: "QC vegetation data"
author: "Nathan Roe"
date: "2024-12-11"
output: html_document
---

**About**

`ecositer` has a category of functions used for QC, all of which use a "QC_" prefix. Vegetation data stored in NASIS requires QC. There are numerous issues that exist caused by different data vintages, data origins, varying data quality, duplication caused by KSSL lab pedons, outdated plant taxonomies, and more. This vignette discusses a best practice methodology for QC NASIS vegetation data. 

You should start your QC process by removing data at the coarsest level that meets your objectives. This will ensure that you do not spend time fixing obscure edge cases that are ultimately removed. 

To begin with, we are going to build our dataset:
```{r}
D104_veg <- ecositer::create_veg_df(from = "web_report",
                                 ecositeid = "F022AD104CA")
```

**Choose the best vegetation plot for site**

Sites may have multiple vegetation plots with differing data qualities. If there are multiple vegetation plots associated with a site, `QC_best_vegplot_for_site()` chooses the vegetation plot with the most records.

```{r}
D104_veg_best <- ecositer::QC_best_vegplot_for_site(veg_df = D104_veg)
```

To look at all the sites with multiple vegetation plots:
```{r}
D104_multi_veg <- D104_veg |> dplyr::group_by(siteiid) |> 
  dplyr::summarise(unique_vegplots = dplyr::n_distinct(vegplotiid)) |> 
  dplyr::mutate(has_multiple_vegplots = unique_vegplots > 1) |> 
  dplyr::filter(has_multiple_vegplots == TRUE)

head(D104_multi_veg)
```


**Creating an aggregated abundance column**

Now, we are going to create an aggregated abundance column called, "pct_cover". There are four columns where abundance data could be stored: akstratumcoverclasspct, speciescancovpct, speciescomppct, understorygrcovpct. A couple of functions will be used to inspect what abundance columns are used in your dataset and whether any records in your dataset have multiple abundance columns used. 

`QC_aggregate_abundance()` aggregates the four abundance columns into the new "pct_cover" column. If any records have multiple abundance columns used,
this function will average them and issue a warning or error depending on the `fail_on_dup` argument. For details on this function - `?QC_aggregate_abundance()`.  
```{r, echo=FALSE}
D104_veg_best[69,18] <- 5
```

```{r}
D104_veg_agg <- ecositer::QC_aggregate_abundance(veg_df = D104_veg_best)
```

Note the two warnings above. As mentioned in the warning, `QC_find_multiple_abundance()` will allow you to inspect which records have abundance data in multiple columns. These should be inspected to determine why multiple abundance columns are used. The other warning shows that two total abundance columns were used in this dataset. 
```{r}
abund_dups <- ecositer::QC_find_multiple_abundance(veg_df = D104_veg)
abund_dups
```

**Assign minimum criteria for vegetation data quality**

Now that we have an aggregated abundance column, we can begin with our coarse filter of the data. First we will use `QC_veg_completeness()` to describe the completeness and quality of vegplot observations. 
```{r}
ecositer::QC_veg_completeness(veg_df = D104_veg_agg) |> head()
```

Next, we can remove plots that do not meet a minimum threshold for data quality and completeness. The threshold should be specific to your data set and study area. 
```{r}
D104_veg_agg <- ecositer::QC_completeness_criteria(veg_df = D104_veg_agg,
                                   min_unique_species = 5,
                                   min_perc_to_species = 60,
                                   min_perc_with_abund = 80)
```

**Update taxonomy using USDA PLANTS**

```{r}
D104_veg_agg <- ecositer::QC_update_taxonomy(veg_df = D104_veg_agg)
```


